Dissecting the interaction between skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights

### 2022-11-18 manuscript/manuscript.Rmd
addressed several minor comments from Reviewer #2.

The morphological variability of Acheulean handaxes has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019]. Despite the recurrent narrative emphasizing the homogeneity and longevity of handaxe assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011]. More specifically, a complex suite of interconnecting factors have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], time budgets [@schillinger2014], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021]. From this extensive list, knapper skill levels and mental templates have been repeatedly mentioned and discussed in the now extensive corpus of handaxe studies, and Boxgrove handaxes have been one of the most studied assemblages from these two angles. Of particular attention here are the experimental works conducted by Stout et al. [-@stout2014] focusing on inferring high knapping skill level and Garcia-Medrano et al. [-@garcía-medrano2019] identifying the mental template of the Boxgrove assemblage. Our paper combines these two perspectives and provides novel insights to the same archaeological assemblage by comparing it with experimentally made handaxes.

In its classical definition, the term mental template indicates that the "idea of the proper form of an object exists in the mind of the maker, and when this idea is expressed in tangible form in raw material, an artifact results" [@deetz1967: 45]. This concept lies at the very foundation of the cultural-historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial-temporal framework. Early researchers, whether explicitly or implicitly, often endorsed this conceptual framework and actively applied it in the typological analysis of handaxes at the regional level [@roe1969; @wenban-smith2000; @wenban-smith2004]. Combined with the production of large flakes, the emergence of mental templates (or "imposed form") has been recognized as a major technological innovation of the Acheulean compared with the Oldowan [@isaac1986]. For a decade or so, this concept has been less frequently used, since it was criticized for a) its normative and static assumption [@lyman2004], b) ignorance of other competing factors such as raw material constraints [@white1995], and c) the lack of rigorous studies of its corresponding cognitive processes. To avoid the historical baggage associated with this controversial term, some researchers developed alternative frameworks such as "design imperatives" derived from utilitarian and ergonomic principles, which refers to a set of minimum features shared by all handaxes including their glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018].

Until recently, researchers have actively addressed the above-mentioned critiques and reconceptualized the mental template in the study of handaxe morphology. Regarding the normative and static assumptions, Hutchence and Scott [-@hutchence2021], for example, leveraged the theory of "community of practice" [@wenger1998] to explain the stability of Boxgrove handaxe design across multiple generations, especially how the social norms behind the consolidated material expressions were developed and negotiated by individuals in a group who have a shared history of learning. They further emphasized that emergent actions of individual knappers also contribute greatly to the shape of Boxgrove handaxes but they were simultaneously constrained by the imposition of social norms. This view also somewhat echoes the "individualized memic construct" proposed by McNabb et al. [-@mcnabb2004], which tries to provide a more balanced perspective incorporating both individual agency and social learning. As for the critique towards confounding factors explaining morphological variability, raw material is often treated as an important variable to be controlled at the very beginning of a research design focusing on mental templates. This is best exemplified by an experimental study of García-Medrano et al. [-@garcía-medrano2019], where they carefully chose experimental nodules mirroring those found in the Boxgrove archaeological assemblage in composition, size, and shape. In terms of the cognitive mechanisms behind mental templates, Ho and colleagues [-@ho2022] recently developed a series of navigation experiments demonstrating the externalization of the planning process to simple geometric representations instead of a complete representation of the given task, featuring both the efficiency and flexibility given the limited cognitive resources. Their experimental design has the potential to be transferred into a research setting aiming at directly testing the planning of knapping behaviors and elucidating how "mental templates" are constructed and perceived in brains. In short, when exercised with proper caution, the concept of mental templates still has its value in our study of handaxe morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness, elongation, and cross-sectional thinning among other things.

@@

Given the number of variables involved in this study, we used principal component analysis (PCA) to reduce the dimension and identify the possible patterns in this morphometric data set, which is one of the most used techniques in similar studies [@iovita2011; @shipton2015; @stout2014; @garcía-medrano2020a; @garcía-medrano2020b; @herzlinger2017]. To detect the effect of training on novices' performance as compared with archaeological samples and handaxe made by experts, we also compare the corresponding metrics built on PCA across different training periods and across all groups using the Games-Howell nonparametric post-hoc test, which does not rely on the assumptions of equal sample sizes and equal variance. This study adheres to the principles of reproducibility and data transparency of archaeological research by depositing all the codes and data sets involved in an open-access online repository [@marwick2017], which can be accessed through the author's Github (<https://github.com/Raylc/Boxgrove-Exp>).

@@

Our analysis suggested that the first two components already explain 77.2% of the variation for the entire morphometric data set composed of 19 variables (**Figure** \@ref(fig:Screeplot)), which is a rather reasonable variance ratio to avoid overfitting. Variable loadings (**Table** \@ref(tab:tab1)) indicate that the first principal component (PC1) captures overall cross-sectional thickness. It is positively correlated with all thickness measurements while negatively correlated with all other measurements. A higher PC1 value thus indicates a thicker handaxe, and vice versa. The second principal component (PC2) tracks elongation and pointedness, as indicated by a positive covariance of maximum length and bottom width/thickness. As PC2 increases, a handaxe will be relatively longer and more convergent from the broad base to the tip.. Thus, PC1 corresponds to cross-sectional thinning and PC2 to overall shape variation.

@@

```{r tab1, echo=FALSE, message=FALSE}
library(tidyverse)
df1<-read.csv(here::here("data/Table1.csv"), fileEncoding = 'UTF-8-BOM')
kableExtra::kbl(df1, booktabs = T, caption = "Variable loadings for the first two principal components") %>%
  kableExtra::kable_styling(position = "left", full_width = TRUE, latex_options = c("striped", "HOLD_position"), font_size = 8)%>%
kableExtra::row_spec(0, bold = T)

@@

We extracted the PC1 and PC2 values of individual handaxes and compared them between different groups. More specifically, the novice group was divided into three sub-groups based on their training stages as specified in the method section. As such, we found that for PC1 values (**Figure** \@ref(fig:PCA1)), the only two group comparisons that are **not** statistically significant are the one between Boxgrove and Expert ($t=-1.65$, $p>0.05$) and the one between Early training and Late training stages ($t=-0.649$, $p>0.05$), which at least partially confirms our visual observation of the general PCA scatter plot. Likewise, for PC2 values (**Figure** \@ref(fig:PCA2)), the group comparison between the Early training and Late stages again is not statistically significant ($t=0.333$, $p>0.05$). An unexpected result is that the mean PC2 value difference between the Pre-training group and Boxgrove is also not statistically significant ($t=-0.818$, $p>0.05$).

>>>>

The morphological variability of Acheulean handaxes has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019]. Despite the recurrent narrative emphasizing the homogeneity and longevity of handaxe assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011]. More specifically, a complex suite of interconnecting factors have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016; @mcnabb2015], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @kohn1999; @white2018; @key2017], reduction method/intensity [@shipton2009; @shipton2015], learning processes [@lycett2016; @kempe2012], time budgets [@schillinger2014], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021]. From this extensive list, knapper skill levels and mental templates have been repeatedly mentioned and discussed in the now extensive corpus of handaxe studies, and Boxgrove handaxes have been one of the most studied assemblages from these two angles. Of particular attention here are the experimental works conducted by Stout et al. [-@stout2014] focusing on inferring high knapping skill level and Garcia-Medrano et al. [-@garcía-medrano2019] identifying the mental template of the Boxgrove assemblage. Our paper combines these two perspectives and provides novel insights to the same archaeological assemblage by comparing it with experimentally made handaxes.

In its classical definition, the term mental template indicates that the "idea of the proper form of an object exists in the mind of the maker, and when this idea is expressed in tangible form in raw material, an artifact results" [@deetz1967: 45]. This concept lies at the very foundation of the cultural-historical approach in that the identification of archaeological cultures is based on the existence of distinct mental templates in a given spatial-temporal framework. Early researchers, whether explicitly or implicitly, often endorsed this conceptual framework and actively applied it in the typological analysis of handaxes at the regional level [@roe1969; @wenban-smith2000; @wenban-smith2004]. Combined with the production of large flakes, the emergence of mental templates (or "imposed form") has been recognized as a major technological innovation of the Acheulean compared with the Oldowan [@isaac1986]. For a decade or so, this concept has been less frequently used, since it was criticized for a) its normative and static assumption [@lyman2004], and b) ignorance of other competing factors such as raw material constraints [@white1995]. To avoid the historical baggage associated with this controversial term, some researchers developed alternative frameworks such as "design imperatives" derived from utilitarian and ergonomic principles, which refers to a set of minimum features shared by all handaxes including their glob-butt, forward extension, support for the working edge, lateral extension, thickness adjustment, and skewness [@gowlett2006; @wynn2018].

Until recently, researchers have actively addressed the above-mentioned critiques and reconceptualized the concept of mental template in the study of handaxe morphology. Regarding the normative and static assumptions, Hutchence and Scott [-@hutchence2021], for example, leveraged the theory of "community of practice" [@wenger1998] to explain the stability of Boxgrove handaxe design across multiple generations, especially how the social norms behind the consolidated material expressions were developed and negotiated by individuals in a group who have a shared history of learning. They further emphasized that emergent actions of individual knappers also contribute greatly to the shape of Boxgrove handaxes but they were simultaneously constrained by the imposition of social norms. This view also somewhat echoes the "individualized memic construct" proposed by McNabb et al. [-@mcnabb2004], which highlighted the influence of individual agency that is complementary to the traditionally favored explanation of social learning. As for the critique towards confounding factors explaining morphological variability, raw material is often treated as an important variable to be controlled at the very beginning of a research design focusing on mental templates. This is best exemplified by an experimental study of García-Medrano et al. [-@garcía-medrano2019], where they carefully chose experimental nodules mirroring those found in the Boxgrove archaeological assemblage in composition, size, and shape. In short, when exercised with proper caution, the concept of mental templates still has its value in our study of handaxe morphological variation, which can be further dissected into a series of shape variables corresponding to pointedness, elongation, and cross-sectional thinning among other things.

@@

As the initial step, simple visualization techniques such scatter plots are frequently used to explore the relationships between variables of interest. Given the number of variables involved in this study, we used principal component analysis (PCA) to reduce the dimension and identify the possible patterns in this morphometric data set, which is one of the most used techniques in similar studies [@iovita2011; @shipton2015; @stout2014; @garcía-medrano2020a; @garcía-medrano2020b; @herzlinger2017]. To detect the effect of training on novices' performance as compared with archaeological samples and handaxe made by experts, we also compare the corresponding metrics built on PCA across different training periods and across all groups using the Games-Howell nonparametric post-hoc test. Compared with other nonparametric tests frequently used in archaeological research for multiple group comparison such as Tukey's test, Games-Howell test does not rely on the assumptions of sample normality, and equal sample sizes and equal variance are not necessary conditions to perform this test. The sample size of each compared group can be as low as 6 [@games1976; @sauder2019]. This study adheres to the principles of reproducibility and data transparency of archaeological research by depositing all the codes and data sets involved in an open-access online repository [@marwick2017], which are available as supplementary materials and can be accessed through the author's Github (<https://github.com/Raylc/Boxgrove-Exp>).

@@

Our analysis suggested that the first two components already explain 77.2% of the variation for the entire morphometric data set composed of 19 variables (**Figure** \@ref(fig:Screeplot)), which is a rather reasonable variance ratio to avoid overfitting. Variable loadings (**Table** \@ref(tab:tab1)) indicate that the first principal component (PC1) captures overall cross-sectional thickness. It is positively correlated with all thickness measurements while negatively correlated with all other measurements. A higher PC1 value thus indicates a thicker handaxe, and vice versa. The second principal component (PC2) tracks elongation and pointedness, as indicated by a positive covariance of maximum length and bottom width/thickness. As PC2 increases, a handaxe will be relatively longer and more convergent from the broad base to the tip. Thus, PC1 corresponds to cross-sectional thinning and PC2 to overall shape variation.

@@

```{r tab1, echo=FALSE, message=FALSE}
library(tidyverse)
df1<-read.csv(here::here("data/Table1.csv"), fileEncoding = 'UTF-8-BOM')
df1<- df1 %>% 
  mutate_if(is.numeric, round, digits = 4) %>% 
  rename(Variables = X)
kableExtra::kbl(df1, booktabs = T, caption = "Variable loadings for the first two principal components.") %>%
  kableExtra::kable_styling(position = "left", full_width = TRUE, latex_options = c("striped", "HOLD_position"), font_size = 8)%>%
kableExtra::row_spec(0, bold = T)

@@

We extracted the PC1 and PC2 values of individual handaxes and compared them between different groups, where the novice group was divided into three sub-groups based on their training stages as specified in the method section. As such, we found that for PC1 values (**Figure** \@ref(fig:PCA1)), the only two group comparisons that are **not** statistically significant are the one between Boxgrove and Expert ($t=-1.65$, $p>0.05$) and the one between Early training and Late training stages ($t=-0.649$, $p>0.05$), which at least partially confirms our visual observation of the general PCA scatter plot. Likewise, for PC2 values (**Figure** \@ref(fig:PCA2)), the group comparison between the Early training and Late stages again is not statistically significant ($t=0.333$, $p>0.05$). An unexpected result is that the mean PC2 value difference between the Pre-training group and Boxgrove is also not statistically significant ($t=-0.818$, $p>0.05$).
